Invasive lionfish present region-wide variation of allometric growth in the Western Atlantic

### 2018-04-20 docs/Manuscript.Rmd
Introduction changes, adding paper folder with some references

A substantial amount of research has been done to describe lionfish feeding ecology in North Carolina \citep{muoz_2011}, the Bahamas \citep{morris_2009,cote_2013}, Northern Gulf of Mexico \citep{dahl_2014}, Mexican Caribbean \citep{valdezmoreno_2012,villaseorderbez_2014}, Belize \citep{hackerott_2017}, and Costa Rica \citep{sandel_2015}. \citet{peake_2018} show that invasive lionfish prey on at least 167 different species across on the tropical and temperate North Atlantic. Their feeding behavior and high consumption rates can reduce recruitment and population sizes of native reef-fish species, and can further endanger reef fish \citep{albins_2008, green_2012,rocha_2015}. (However, see \citet{hackerott_2017} for a case where there was no evidence that lionfish affected the density, richness, or community composition of prey fishes). Major efforts have been made to understand the possible impacts of the invasion by keeping track of its range through time \citep{schofield_2009,schofield_2010} and predicting invasion ranges under future climates \citep{grieve_2016}. By combining information from these disciplines, researchers have been able to predict the trophic impacts of lionfish \citep{ariasgonzalez_2011}, which can then be translated into ecosystem-level and economic impacts.

Seeking to reduce lionfish densities, governments and non-profit organizations have promoted removal programs and incentivized its consumption \citep{chin_2016}. In some cases, these have shown to significantly reduce --but not quite eliminate-- lionfish abundances at local scales \citep{sandel_2015,chin_2016,deleon_2013}. Complete eradication of lionfish through fishing is unlikely because of their rapid recovery rates and ongoing recruitment to shallow-water areas from their persistent populations in mesophotic coral ecosystems \citep{barbour_2011,andradibrown_2017}. However, further incentivizing its consumption might create a demand big enough to promote and sustain a stable fishery \citep{chin_2016}, which can reduce local abundances and control the invasion while providing alternative livelihoods.

The feasibility of lionfish removal programs has been extensively evaluated through field observations \citep{usseglio_2017,sandel_2015,chin_2016,deleon_2013} and empirical modeling \citep{barbour_2011,morris_2011,johnston_2015}. The latter approach models changes in biomass in response to changes in mortality (*i.e.* culling). In this case, biomass represents the sum of all fish's individual weight. The individual weight of an organism can be estimated from its length using the allometric growth equation. Parameters for this equation exist for North Carolina, Northern and Southern Gulf of Mexico, the Southern Mexican Caribbean, Bahamas, Little Cayman, Jamaica, Bonaire, Puerto Rico, and Costa Rica, but remain unavailable for the central Mexican Caribbean \citep{barbour_2011,fogg_2013,dahl_2014,aguilarperera_2016,sabidoitza_2016,darling_2011,edwards_2014,chin_2016,deleon_2013,toledohernndez_2014,sandel_2015}. The length-weight of a species can vary across regions as a response to biotic (*e.g.* local food availability) and abiotic (*e.g.* water temperature) conditions \citep{johnson_2016}. Thus, when using biomass-informed models or estimating weights from length observations, it is important to use site-specific parameters. This is especially important when research involves identifying the total biomass available for harvest by fishers or the efficacy of lionfish removals \citep{barbour_2011,morris_2011,johnston_2015,chin_2016}.

This study reviews lionfish allometric growth throughout the invasion range, and provides a new pair of parameters specific to the central Meixcan Caribbean. Results suggest there are important regional-scale variations in allometric growth patterns of lionfish. The observed differences highlight the importance of using site-specific parameters, especially when informing invasion management strategies.

<!-- make this last paragraph a bit longer to talk about literature on spatial differences in fish populations -->

>>>>

A substantial amount of research has been done to describe lionfish feeding ecology in North Carolina \citep{muoz_2011}, the Bahamas \citep{morris_2009,cote_2013}, Northern Gulf of Mexico \citep{dahl_2014}, Mexican Caribbean \citep{valdezmoreno_2012,villaseorderbez_2014}, Belize \citep{hackerott_2017}, and Costa Rica \citep{sandel_2015}. \citet{peake_2018} show that invasive lionfish prey on at least 167 different species across the tropical and temperate North Atlantic. Their feeding behavior and high consumption rates can reduce recruitment and population sizes of native reef-fish species, and can further endanger reef fish \citep{albins_2008, green_2012,rocha_2015}. (However, see \citet{hackerott_2017} for a case where there was no evidence that lionfish affected the density, richness, or community composition of prey fishes). Major efforts have been made to understand the possible impacts of the invasion by keeping track of its range through time \citep{schofield_2009,schofield_2010} and predicting invasion ranges under future climates \citep{grieve_2016}. By combining information from these disciplines, researchers have been able to predict the trophic impacts of lionfish \citep{ariasgonzalez_2011}, which can then be translated into ecosystem-level and economic impacts.

Governments and non-profit organizations have sought to reduce lionfish densities through removal programs and incentivizing its consumption \citep{chin_2016}. In some cases, these have shown to significantly reduce --but not quite eliminate-- lionfish abundances at local scales \citep{sandel_2015,chin_2016,deleon_2013}. In addition, culling programs can help stabilize or grow native prey fish populations (Cote et al. 2014 - https://peerj.com/preprints/398v1.pdf). Complete eradication of lionfish through fishing is unlikely because of their rapid recovery rates and ongoing recruitment to shallow-water areas from their persistent populations in mesophotic coral ecosystems \citep{barbour_2011,andradibrown_2017}. However, promoting its consumption might create a level of demand capable of sustaining a stable fishery, which can help control invasive populations while providing alternative livelihoods \citep{chin_2016}.

The feasibility **of establishing fisheries** (I think this is what we should focus on as our selling point - may mean you have fewer papers to cite but that's okay) through lionfish removal programs has been extensively evaluated through field observations and empirical modeling \citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,chin_2016,usseglio_2017}. One contributing factor to the success of many removal programs is the sedentary nature of adult lionfish (Jud & Layman 2012 - http://craiglaymanlab.com/wp-content/uploads/Jud-and-Layman-2012-JEMBE.pdf). Culling programs are effective in reducing adult populations largely because lionfish exhibit high levels of site fidelity and rarely leave their home range in most cases (Fishelson 1997, Kochzius & Blohm 2005, Cote et al. 2014). As a result of this sedentary behavior, lionfish are also likely to exhibit high levels of spatial variation in important life history characterstics such as growth or natural mortality rate (Hutchinson 2008, Wilson et al. 2012). The importance of considering spatial heterogeneity is well-documented in terms of assessing and managing sedentary species (Gunderson et al. 2008, Guan et al. 2013), and such variation should be accounted for when evaluating the feasibility of establishing lionfish fisheries as well.

Empirical modeling efforts examining the feasibility of establishing fisheries for lionfish involve modeling changes in biomass in response to changes in mortality (*i.e.* culling - **cite some of the papers from above paragraph**). In this case, biomass represents the sum of all individual fish weights. The individual weight (*W*) of an organism can be estimated from its length using the allometric growth equation, $W = aL^b$, where *L* is an organism's length and *a* and *b* are constants **(I know it sems a little weird to put this in the intro, but since it's the whole crux of the paper I think it's okay - more details still follow later and it didn't take much space...)**. This length-weight relationship for any given species is an essential component of biomass-based modeling, and can vary across regions as a response to biotic (*e.g.* local food availability) and abiotic (*e.g.* water temperature) conditions \citep{johnson_2016}. Site-specific parameters are therefore necessary in order to obtain accurate biomass estimates in a given region, especially when research involves identifying the total biomass available for harvest by fishers \citep{barbour_2011,morris_2011,johnston_2015,chin_2016}. Equation parameters for lionfish exist for North Carolina, Northern and Southern Gulf of Mexico, the Southern Mexican Caribbean, Bahamas, Little Cayman, Jamaica, Bonaire, Puerto Rico, and Costa Rica, but remain unavailable for the central Mexican Caribbean \citep{barbour_2011,fogg_2013,dahl_2014,aguilarperera_2016,sabidoitza_2016,darling_2011,edwards_2014,chin_2016,deleon_2013,toledohernndez_2014,sandel_2015}.

Here, we review lionfish allometric growth parameters throughout its invasion range and examine spatial differences in the equation constants. We also provide a new pair of parameters specific to the central Meixcan Caribbean. Results suggest there are important regional-scale variations in allometric growth patterns of lionfish, and these differences highlight the importance of using site-specific parameters when developing or informing invasion management strategies.

### 2018-04-20 docs/Manuscript.tex
Added Sean's name to list of authors
Introduction changes, adding paper folder with some references

\author[1]{Juan Carlos Villaseñor-Derbez}
\affil[1]{Bren School of Environmental Sciences and Management, University of
California Santa Barbara, Santa Barbara, California, U.S.}
\corrauthor[1]{Juan Carlos Villaseñor-Derbez}{jvillasenor@bren.ucsb.edu}

@@

A substantial amount of research has been done to describe lionfish
feeding ecology in North Carolina \citep{muoz_2011}, the Bahamas
\citep{morris_2009,cote_2013}, Northern Gulf of Mexico
\citep{dahl_2014}, Mexican Caribbean
\citep{valdezmoreno_2012,villaseorderbez_2014}, Belize
\citep{hackerott_2017}, and Costa Rica \citep{sandel_2015}.
\citet{peake_2018} show that invasive lionfish prey on at least 167
different species across on the tropical and temperate North Atlantic.
Their feeding behavior and high consumption rates can reduce recruitment
and population sizes of native reef-fish species, and can further
endanger reef fish \citep{albins_2008, green_2012,rocha_2015}. (However,
see \citet{hackerott_2017} for a case where there was no evidence that
lionfish affected the density, richness, or community composition of
prey fishes). Major efforts have been made to understand the possible
impacts of the invasion by keeping track of its range through time
\citep{schofield_2009,schofield_2010} and predicting invasion ranges
under future climates \citep{grieve_2016}. By combining information from
these disciplines, researchers have been able to predict the trophic
impacts of lionfish \citep{ariasgonzalez_2011}, which can then be
translated into ecosystem-level and economic impacts.

Seeking to reduce lionfish densities, governments and non-profit
organizations have promoted removal programs and incentivized its
consumption \citep{chin_2016}. In some cases, these have shown to
significantly reduce --but not quite eliminate-- lionfish abundances at
local scales \citep{sandel_2015,chin_2016,deleon_2013}. Complete
eradication of lionfish through fishing is unlikely because of their
rapid recovery rates and ongoing recruitment to shallow-water areas from
their persistent populations in mesophotic coral ecosystems
\citep{barbour_2011,andradibrown_2017}. However, further incentivizing
its consumption might create a demand big enough to promote and sustain
a stable fishery \citep{chin_2016}, which can reduce local abundances
and control the invasion while providing alternative livelihoods.

The feasibility of lionfish removal programs has been extensively
evaluated through field observations
\citep{usseglio_2017,sandel_2015,chin_2016,deleon_2013} and empirical
modeling \citep{barbour_2011,morris_2011,johnston_2015}. The latter
approach models changes in biomass in response to changes in mortality
(\emph{i.e.} culling). In this case, biomass represents the sum of all
fish's individual weight. The individual weight of an organism can be
estimated from its length using the allometric growth equation.
Parameters for this equation exist for North Carolina, Northern and
Southern Gulf of Mexico, the Southern Mexican Caribbean, Bahamas, Little
Cayman, Jamaica, Bonaire, Puerto Rico, and Costa Rica, but remain
unavailable for the central Mexican Caribbean
\citep{barbour_2011,fogg_2013,dahl_2014,aguilarperera_2016,sabidoitza_2016,darling_2011,edwards_2014,chin_2016,deleon_2013,toledohernndez_2014,sandel_2015}.
The length-weight of a species can vary across regions as a response to
biotic (\emph{e.g.} local food availability) and abiotic (\emph{e.g.}
water temperature) conditions \citep{johnson_2016}. Thus, when using
biomass-informed models or estimating weights from length observations,
it is important to use site-specific parameters. This is especially
important when research involves identifying the total biomass available
for harvest by fishers or the efficacy of lionfish removals
\citep{barbour_2011,morris_2011,johnston_2015,chin_2016}.

This study reviews lionfish allometric growth throughout the invasion
range, and provides a new pair of parameters specific to the central
Meixcan Caribbean. Results suggest there are important regional-scale
variations in allometric growth patterns of lionfish. The observed
differences highlight the importance of using site-specific parameters,
especially when informing invasion management strategies.

>>>>

\author[1]{Juan Carlos VillaseÃƒÂ±or-Derbez, Sean Fitzgerald}
\affil[1]{Bren School of Environmental Sciences and Management, University of
California Santa Barbara, Santa Barbara, California, U.S.}
\corrauthor[1]{Juan Carlos VillaseÃƒÂ±or-Derbez, Sean Fitzgerald}{jvillasenor@bren.ucsb.edu}

@@

A substantial amount of research has been done to describe lionfish
feeding ecology in North Carolina \citep{muoz_2011}, the Bahamas
\citep{morris_2009,cote_2013}, Northern Gulf of Mexico
\citep{dahl_2014}, Mexican Caribbean
\citep{valdezmoreno_2012,villaseorderbez_2014}, Belize
\citep{hackerott_2017}, and Costa Rica \citep{sandel_2015}.
\citet{peake_2018} show that invasive lionfish prey on at least 167
different species across the tropical and temperate North Atlantic.
Their feeding behavior and high consumption rates can reduce recruitment
and population sizes of native reef-fish species, and can further
endanger reef fish \citep{albins_2008, green_2012,rocha_2015}. (However,
see \citet{hackerott_2017} for a case where there was no evidence that
lionfish affected the density, richness, or community composition of
prey fishes). Major efforts have been made to understand the possible
impacts of the invasion by keeping track of its range through time
\citep{schofield_2009,schofield_2010} and predicting invasion ranges
under future climates \citep{grieve_2016}. By combining information from
these disciplines, researchers have been able to predict the trophic
impacts of lionfish \citep{ariasgonzalez_2011}, which can then be
translated into ecosystem-level and economic impacts.

Governments and non-profit organizations have sought to reduce lionfish
densities through removal programs and incentivizing its consumption
\citep{chin_2016}. In some cases, these have shown to significantly
reduce --but not quite eliminate-- lionfish abundances at local scales
\citep{sandel_2015,chin_2016,deleon_2013}. In addition, culling programs
can help stabilize or grow native prey fish populations (Cote et al.
2014 - \url{https://peerj.com/preprints/398v1.pdf}). Complete
eradication of lionfish through fishing is unlikely because of their
rapid recovery rates and ongoing recruitment to shallow-water areas from
their persistent populations in mesophotic coral ecosystems
\citep{barbour_2011,andradibrown_2017}. However, promoting its
consumption might create a level of demand capable of sustaining a
stable fishery, which can help control invasive populations while
providing alternative livelihoods \citep{chin_2016}.

The feasibility \textbf{of establishing fisheries} (I think this is what
we should focus on as our selling point - may mean you have fewer papers
to cite but that's okay) through lionfish removal programs has been
extensively evaluated through field observations and empirical modeling
\citep{barbour_2011,morris_2011,deleon_2013,johnston_2015,sandel_2015,chin_2016,usseglio_2017}.
One contributing factor to the success of many removal programs is the
sedentary nature of adult lionfish (Jud \& Layman 2012 -
\url{http://craiglaymanlab.com/wp-content/uploads/Jud-and-Layman-2012-JEMBE.pdf}).
Culling programs are effective in reducing adult populations largely
because lionfish exhibit high levels of site fidelity and rarely leave
their home range in most cases (Fishelson 1997, Kochzius \& Blohm 2005,
Cote et al. 2014). As a result of this sedentary behavior, lionfish are
also likely to exhibit high levels of spatial variation in important
life history characterstics such as growth or natural mortality rate
(Hutchinson 2008, Wilson et al. 2012). The importance of considering
spatial heterogeneity is well-documented in terms of assessing and
managing sedentary species (Gunderson et al. 2008, Guan et al. 2013),
and such variation should be accounted for when evaluating the
feasibility of establishing lionfish fisheries as well.

Empirical modeling efforts examining the feasibility of establishing
fisheries for lionfish involve modeling changes in biomass in response
to changes in mortality (\emph{i.e.} culling - \textbf{cite some of the
papers from above paragraph}). In this case, biomass represents the sum
of all individual fish weights. The individual weight (\emph{W}) of an
organism can be estimated from its length using the allometric growth
equation, \(W = aL^b\), where \emph{L} is an organism's length and
\emph{a} and \emph{b} are constants \textbf{(I know it sems a little
weird to put this in the intro, but since it's the whole crux of the
paper I think it's okay - more details still follow later and it didn't
take much space\ldots{})}. This length-weight relationship for any given
species is an essential component of biomass-based modeling, and can
vary across regions as a response to biotic (\emph{e.g.} local food
availability) and abiotic (\emph{e.g.} water temperature) conditions
\citep{johnson_2016}. Site-specific parameters are therefore necessary
in order to obtain accurate biomass estimates in a given region,
especially when research involves identifying the total biomass
available for harvest by fishers
\citep{barbour_2011,morris_2011,johnston_2015,chin_2016}. Equation
parameters for lionfish exist for North Carolina, Northern and Southern
Gulf of Mexico, the Southern Mexican Caribbean, Bahamas, Little Cayman,
Jamaica, Bonaire, Puerto Rico, and Costa Rica, but remain unavailable
for the central Mexican Caribbean
\citep{barbour_2011,fogg_2013,dahl_2014,aguilarperera_2016,sabidoitza_2016,darling_2011,edwards_2014,chin_2016,deleon_2013,toledohernndez_2014,sandel_2015}.

Here, we review lionfish allometric growth parameters throughout its
invasion range and examine spatial differences in the equation
constants. We also provide a new pair of parameters specific to the
central Meixcan Caribbean. Results suggest there are important
regional-scale variations in allometric growth patterns of lionfish, and
these differences highlight the importance of using site-specific
parameters when developing or informing invasion management strategies.
